Pattners of stress tolerance versus resistance in genome-wide expression data

### 2014-07-15 doc/ApTranscriptome_manuscript.Rmd
Updates to ms

---
title: "Thermal reactionome of a temperate ant species"
author: "John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn,
  Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan"
csl: ecology.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
  html_document: null
bibliography: ApTranscriptome.bib
---

@@

The study of whole organism gene expression (e.g. transcriptomics) has rapidly increased due to improved sequencing technology (RNAseq). To date, RNAseq studies have largely focused on differential expression across few environments [@teets2012; @pespeni2013; @barshis2013; @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach. However, many ecologically-relevant aspects of the environment such as temperature vary continuously in nature. Variation in the phenotype of an organism in response to such a continuous predictor (i.e. phenotypic plasticity) is referred to as a reaction norm. A reaction norm can differ not only in mean values, but also in the shape of the curve as is often found for thermal reaction norms [@hoffmann2013; @murren2014]. An analysis of phenotypic reaction norms found that differences in the shape were greater than differences in mean values in comparisons among species and populations [@murren2014]. The authors concluded that many patterns of biologically-relevant plasticity are likely to be missed in studies with few environments. Similarly, gene expression studies with few environments are likely to miss many biologically-relevant changes in gene expression. To overcome this limitation, we use a novel method to characterize the reaction norm of all genes in the transcriptome (e.g. the reactionome) in response to thermal stress. Specifically, we measure gene expression across a thermal gradient and identify thermally-responsive genes using a regression approach.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
options(scipen = 1, digits = 3)
load("../ms.RData")
library(data.table)
library(pander)
library(ggplot2)
library(topGO)
```

@@

We identified `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` genes with overall model fits of *P* < 0.05, retaining `r prettyNum(length(which(RxNpval$padj < 0.05)), big.mark = ",", scientific = FALSE)` after adjusting *P* values for false discovery rate. Of these genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` significantly change expression with temperature including `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` that have a species x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not thermally-responsive in one species (*NotResp*) differed significantly between the species ($\chi^2$ = `r chi1$statistic`, P < `r max(chi1$p.value, 0.001)`; Table 1). The density function for the temperature of maximum expression (Fig. 1) showed that the majority of responsive genes were in the *Low* category for both species, with *A. picea* having more *Low* genes than *A. carolinensis*. More genes had *Intermediate* expression in *A. carolinensis* than in *A. picea*, while the two species had a similar number of *High* genes (Fig. 1, Table 1).

```{r expression_type_table, echo=FALSE, results='asis'}
pandoc.table(tt3, style="rmarkdown", split.tables = Inf, caption = "Number of thermally-responsive genes in each expression category for each species. *A. carolinensis* in rows and *A. picea* in columns.")
```

```{r exp_plot, echo=FALSE, warning = FALSE, fig.cap = "Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. carolinensis* (red) species"}
ggplot(Ap.df, aes(x=max.val, fill=colony, colour = colony)) +
  geom_density(alpha = 0.2, position = "identity") +
  scale_y_continuous(name = "Density") +
  scale_x_continuous(name = "Temperature of maximum expression")
```

>>>>

---
title: "Pattners of stress tolerance versus resistance in genome-wide expression data
  of sympatric ant species"
author: "John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn,
  Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan"
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
  html_document: null
  word_document: default
bibliography: ApTranscriptome.bib
---

@@

The study of whole organism gene expression (e.g. transcriptomics) has rapidly increased due to improved RNA sequencing (RNAseq) technology. To date, RNAseq studies have largely focused on differential expression across few environments [@teets2012; @pespeni2013; @barshis2013; @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach. However, many ecologically-relevant aspects of the environment such as temperature vary continuously in nature. Variation in the phenotype of an organism in response to such a continuous predictor (i.e. phenotypic plasticity) is referred to as a reaction norm. A reaction norm can differ not only in mean values, but also in the shape of the curve as is often found for thermal reaction norms [@hoffmann2013; @murren2014]. An analysis of phenotypic reaction norms found that differences in the shape were greater than differences in mean values in comparisons among species and populations [@murren2014]. Murren et al [-@murren2014] concluded that many patterns of biologically-relevant plasticity are likely to be missed in studies with few environments. Similarly, gene expression studies with few environments are likely to miss many biologically-relevant changes in gene expression.

For evolutionary ecology studies, the goal of RNAseq studies is often to identify the genetic basis of environmental adaptation [cite..] and provide insight on evolutionary potential. However, the results of RNAseq provides, lists of differentially-expressed genes, are prone to false positives [cite...] and miss genes with small expression changes that have large phenotypic effects [cite...], thus limiting their usefullness in this context. Genome-wide association (GWA) studies have faced a similar problem where the phenotypic effect of the few statistically-significant genes is over-estimated while many causal genes of small effect are missed [@beavis; @larson2013]. Both our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are undermined by these issues. Thus, plant and animal breeders have largely abandoned GWA in favor of genomic prediction. Analog

In this study, we apply a novel method to characterize the reaction-norm of all genes in the transcriptome (e.g. the reactionome) in response to thermal stress to overcome the above limitations. Specifically, we measure gene expression across a thermal gradient and identify thermally-responsive genes using a regression approach. Instead of focusing on individual genes or groups of genes with differential expression, we compare overall patterns in the reactionomes between closely-related species to gain insight into how spatially-divergent selection has changed gene expression patterns.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
options(scipen = 1, digits = 3)
load("../ms.RData")
library(data.table)
library(pander)
library(ggplot2)
library(topGO)
library(RColorBrewer)
```

@@

```{r mosaic_plot, echo=FALSE, results='hide'}
tiff("mosaic_plot.tiff")
mosaic.colors <- brewer.pal(5, "Blues")
par(mar = c(5.1, 3.1, 2.1, 7.5))
mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
labels <- c("Low", "Intermediate", "High", "Bimodal", "NotResp")
text(c(.1,.5,.75,.85,.99), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
text(.5, par("usr")[4]+.1, "A. carolinensis", xpd = TRUE, font = 3)
text(.2, .6, "2,594")
text(.85, .03, "223", cex = .75)
legend(1.05, .9, fill = mosaic.colors, c("Low", "Intermediate", "High", "Bimodal", "NotResp"), xpd=TRUE)
dev.off()
```

![Mosaic plot of the number of thermally-responsive genes by expression type. The width of the columns is proportional to the number of genes for each expression type for *A. carolinensis* while the height of the rows is proportional to the number of genes for each expression type for *A. picea.* The rows are colored by expression type for *A. picea*. The number of genes are labeled in two boxes for scale.](mosaic_plot.tiff)

We identified `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` genes with overall model fits of *P* < 0.05, retaining `r prettyNum(length(which(RxNpval$padj < 0.05)), big.mark = ",", scientific = FALSE)` after adjusting *P* values for false discovery rate. Of these genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` significantly change expression with temperature including `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` that have a species x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not thermally-responsive in one species (*NotResp*) differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Table 1). The density function for the temperature of maximum expression (Fig. 1) showed that the majority of responsive genes were in the *Low* category for both species, with *A. picea* having more *Low* genes than *A. carolinensis*. More genes had *Intermediate* expression in *A. carolinensis* than in *A. picea*, while the two species had a similar number of *High* genes (Fig. 1, Table 1).
